---
title: "graphs"
output: 
  html_document: 
    highlight: monochrome
    theme: journal
date: "2025-07-02"
---

```{r install, echo=FALSE, message=FALSE,warning=FALSE}

#install.packages("readxl")
#install.packages("readr")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("ggplot2")
#install.packages("stringr")
#install.packages("dplyr")
#install.packages("brms")
#install.packages("knitr")
#install.packages("kableExtra")

library(brms)
library(knitr)
library(kableExtra)
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(dplyr)

combined <- readRDS("combined.rds")
```

```{r sib count, echo=FALSE, message=FALSE,warning=FALSE}
sib_yesno <- table(
  factor(
    combined$`Are there any other children who live in your home?`,
    levels = c("Yes", "No")
  )
)

sib_yesno <- as.data.frame(sib_yesno)
colnames(sib_yesno) <- c("Answer", "Count")
```

### Household and Sibling Data

Refers to sibship factors such as sibling identity, ages of children, number of children in a household, etc.
<br>
<br>

```{r sibling count graph, fig.width = 5, fig.height = 5, echo=FALSE, message=FALSE,warning=FALSE}
householdplot <- ggplot(sib_yesno, aes(x = Answer, y = Count)) + 
  geom_bar(stat = "identity", fill = "plum") +
  geom_text(aes(label = Count), vjust = -0.5, family = "Times") +
  labs(title = "How Many Households have Siblings?"
  ) + 
  theme_minimal(base_size = 12, base_family = "Times") +
    theme(plot.title = element_text(hjust = 0.5, size = 10, face = "bold")) 

householdplot
```

```{r sibling identities code, echo=FALSE, message=FALSE,warning=FALSE}
sib_df <- combined %>%
  select(
    parent_ID =  subject_id,
    has_siblings = `Are there any other children who live in your home?`, 
    info = `Please enter the birthday (MM/DD/YYYY) and the sex of the children:`
  )

sib_df <- sib_df %>%
  filter(has_siblings == "Yes", !is.na(info))
```

```{r graph of sibling identity, fig.width = 7, fig.height = 6, echo=FALSE, message=FALSE,warning=FALSE}
setwd("/Volumes/Fas-Phyc-PEB-Lab/experimental_projects/PBS/PBS_Full_Study/CN_Working_Folder/R/Survey")
sibship_factors <- read_xlsx("Sibship_Factors.xlsx")

sibship_factors <- sibship_factors %>%
  filter(Sib_Identity != "n/a", Sib_Identity != "", !is.na(Sib_Identity))

ggplot(sibship_factors, aes(x = Sib_Identity))+
  geom_bar(fill = "lightcoral") +
  geom_text(stat = "count", aes(label = ..count..),vjust = -0.3) +
  labs(
    title = "Sibling Identities",
    x = "Sibling Identity",
    y = "Count"
  ) +
  theme_minimal(base_size = 12, base_family = "Times") + 
      theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) 

```

`Sibling Identity` refers to the sibling's relation to the target child.

```{r sibling count in households, fig.width = 5, fig.height = 8, echo=FALSE, message=FALSE,warning=FALSE}
household_counts <- sibship_factors %>%
  filter(Sib_Birthdate != "n/a") %>%
  group_by(Subject_ID) %>%
  summarise(num_siblings = n(), .groups = "drop")

sib_distribution <- household_counts %>%
  count(num_siblings)

ggplot(sib_distribution, aes(x = as.factor(num_siblings), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = n), vjust = -0.3) +
  labs(
    title = "Sibling Count in Households",
    x = "Number of Siblings",
    y = "Number of Households"
  ) +
  theme_minimal(base_size = 15, base_family = "Times") + 
      theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) 
```

`Sibling count` refers to how many siblings the target child has. In this context, `1` refers to a household of two children.

```{r languages in household + cleaning, echo=FALSE, message=FALSE,warning=FALSE}

lang_other <- combined %>%
  select(subject_id, `Are there any other languages spoken in your home?`) %>%
  rename(`Other Language?` = `Are there any other languages spoken in your home?`) %>%
  mutate(`Other Language?` = tolower(`Other Language?`)) %>%
  filter(!`Other Language?` %in% c("no", "none", "english", "no /youtube french or spanish vidoes"))

colnames(lang_other)[1] <- "Other Language?"
colnames(lang_other)[2] <- "Count"

```


<br>
<br>
### Languages within Households

<br> Out of 74 fully completed observations, 19 households speak another language, that is not English. This is after cleaning and removing any English answers or no answers. `Multilingual` refers to households that speak more than one language group.

```{r graph of other lang spoken, fig.width = 6, fig.height = 8, echo=FALSE, message=FALSE,warning=FALSE}
lang_other <- lang_other %>%
  mutate(Language_Group = case_when(
    grepl(",", `Other Language?`) ~ "Multilingual",  # overrides all
    grepl("spanish", `Other Language?`, ignore.case = TRUE) ~ "Spanish",
    `Other Language?` %in% c("cantonese", "mandarin chinese", "korean", "malyalam", "urdu", "mandarin") ~ "Asian",
    `Other Language?` %in% c("arabic", "farsi", "hebrew") ~ "Middle Eastern",
    `Other Language?` == "russian" ~ "Slavic",
    `Other Language?` == "albanian" ~ "Indo-European",
    `Other Language?` == "asl" ~ "Sign Language",
    TRUE ~ NA_character_
  ))

lang_other %>%
  group_by(Language_Group) %>%
  summarise(Count = sum(Count), .groups = "drop") %>%
  ggplot(aes(x = Language_Group, y = Count)) +
  geom_bar(stat = "identity", fill = "plum", width = 0.7) +
  geom_text(aes(label = Count), vjust = -0.3, family = "Times") +
  labs(
    title = "Other Languages Spoken",
    x = "Language Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 10, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 

```

<br>

### Parent and Child Scores

```{r parent score of overall data, echo=FALSE, message=FALSE,warning=FALSE}
combined$Q64_1 <- as.numeric(combined$Q64_1)
combined$Q38_1 <- as.numeric(combined$Q38_1)
combined$Q39_1 <- as.numeric(combined$Q39_1)
combined$Q54_1 <- as.numeric(combined$Q54_1)
combined$Q65_1 <- as.numeric(combined$Q65_1)

combined$parent_score <- rowSums(combined[, c("Q64_1", "Q54_1", "Q65_1", "Q38_1", "Q39_1")], na.rm = TRUE)
big_text_theme <- theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 14)
  )

combined <- combined %>%
  filter(parent_score > 0)

ggplot(combined, aes(x = parent_score)) +
  geom_histogram(binwidth = 2, fill = "#0F6B99", color = "white", alpha = 0.8) +
  labs(
    title = "Parent Scores",
    x = "Parent Score",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 

```

```{r child score of overall data, echo=FALSE, message=FALSE,warning=FALSE}
combined$Q55_1 <- as.numeric(combined$Q55_1)
combined$Q56_1 <- as.numeric(combined$Q56_1)
combined$Q57_1 <- as.numeric(combined$Q57_1)
combined$Q43_1 <- as.numeric(combined$Q43_1)
combined$Q44_1 <- as.numeric(combined$Q44_1)

combined$child_score <- rowSums(combined[, c("Q55_1", "Q56_1", "Q57_1", "Q43_1", "Q44_1")], na.rm = TRUE)

combined <- combined %>%
  filter(child_score > 0)

ggplot(combined, aes(x = child_score)) +
  geom_histogram(binwidth = 2, fill = "#0F6B99", color = "white", alpha = 0.8) +
  labs(
    title = "Child Scores",
    x = "Child Score",
    y = "Frequency"
  )  +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 
```

<br> <br> <br> <br> <br>
### Box Plots

```{r box plot parent score, echo=FALSE, message=FALSE,warning=FALSE}
ggplot(combined, aes(x = `Are there any other children who live in your home?`, y = parent_score)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "red") +
  labs(
    title = "Parent Scores of Families w/ More Than One Kid",
    y = "Parent Score"
  ) +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 
```

```{r box plot child score, echo=FALSE, message=FALSE,warning=FALSE}
ggplot(combined, aes(x = `Are there any other children who live in your home?`, y = child_score)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "red") +
  labs(
    title = "Child Scores of Families w/ More Than One Kid",
    y = "Child Score"
  ) +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 
```

```{r, echo=FALSE, message=FALSE,warning=FALSE}
ggplot(combined, aes(x = `Are there any other children who live in your home?`, y = Q54_1)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "red") +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) 
```

```{r, echo=FALSE, message=FALSE,warning=FALSE}
combined$`Words Produced` <- as.numeric(combined$`Words Produced`)

ggplot(combined, aes(x = `Are there any other children who live in your home?`, y = `Words Produced`)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") + stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "red") +
  theme_minimal(base_size = 12, base_family = "Times") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  theme_classic(base_size = 14) 
  
```

<br> <br> The CI seen in red shows the uncertainty due to sampling variability, likely because of the small sample size.

<br>
<br>
**\# Analysis**
Because we have both the MB-CDI (words produced) and the child score, we can use a multilevel model to model the two dependent variables. This is done to see if both variables are influenced by the same factors and if the two variables are correlated. 
```{r}

```



<br> **\# IGNORE FOR NOW**

```{r parent score mean w/ & w/o sibs, echo=FALSE, message=FALSE,warning=FALSE}
mean_with_siblings <- mean(combined$parent_score[
  combined$`Are there any other children who live in your home?` == "Yes"
], na.rm = TRUE)

mean_without_siblings <- mean(combined$parent_score[
  combined$`Are there any other children who live in your home?` == "No"
], na.rm = TRUE)

mean_with_siblings
mean_without_siblings
```

```{r child score mean w/ & w/o sibs}
child_mean_with_siblings <- mean(combined$child_score[
  combined$`Are there any other children who live in your home?` == "Yes"
], na.rm = TRUE)

child_mean_without_siblings <- mean(combined$child_score[
  combined$`Are there any other children who live in your home?` == "No"
], na.rm = TRUE)

child_mean_with_siblings
child_mean_without_siblings
```

```{r}
child_mean_lang <- mean(combined$child_score[
  combined$`Are there any other languages spoken in your home?` != c("no","No","English","english")
], na.rm = TRUE)

parent_mean_lang <- mean(combined$parent_score[
  combined$`Are there any other languages spoken in your home?` != c("no","No","English","english")
], na.rm = TRUE)

parent_mean_lang
child_mean_lang
```

```{r}
child_mean_nolang <- mean(combined$child_score[
  combined$`Are there any other languages spoken in your home?` == c("no","No","English","english")
], na.rm = TRUE)

parent_mean_nolang <- mean(combined$parent_score[
  combined$`Are there any other languages spoken in your home?` == c("no","No","English","english")
], na.rm = TRUE)

parent_mean_nolang
child_mean_nolang
```

```{r}
Q54_mean_lang <- mean(combined$Q54_1[
  combined$`Are there any other languages spoken in your home?` != c("no","No","English","english")
], na.rm = TRUE)

Q54_mean_nolang <- mean(combined$Q54_1[
  combined$`Are there any other languages spoken in your home?` == c("no","No","English","english")
], na.rm = TRUE)

Q54_mean_lang
Q54_mean_nolang
```

```{r comparing older/younger target child, echo=FALSE, message=FALSE,warning=FALSE}
range(combined$parent_score, na.rm = TRUE)
mean(c(14, 33))
```
